Title Pending

### 2019-02-04 manuscript/manuscript.tex
edits to discussion

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/04_tree.png}}%
  }
  \par \textbf{Figure 5:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A).
\end{figure}

@@

All of the anxiety disorders share core symptoms including pessimistic inference, second-order conditioning, and sustained avoidance. In this article, we offered a simple computational account of how these aberrations may arise and be maintained. Following the literature in Markov decision processes, we showed how a simple breakdown in expectations about the likelihood of encountering future negative outcomes (as a result of an unpredictable environment or unreliable policy) can effectively backpropagate negative expectations through an agent's internal value map. The result of which yields expectations and behaviors resembling that observed in pathological anxiety. Importantly, we also showed that such a model can accountfor aberrant behavior previously observed in individuals with anxiety in several assays of psychiatric behavior. By no means a complete theory of anxiety, this account helps to at least tie together several core symptoms of anxiety in one account.

This computational account draws upon a longstanding recognition of the importance of perceived control to the development and maintenance of anxiety disorders. Several prominent theories of anxiety in the clinical anxiety focus on the importance of a lack of perceived control, including helplessness-hopelessness theories, perceived efficacy, and the triple vulnerability model. It is worth noting that that these theories differ, among other things, in where a lack of control arises. For learned helplessness, an agent believes to occupy an uncontrollable world; in other words, there is no reliable mapping between action and outcome. For self-efficacy theory, an agent believes to be limited or incapable of exerting beneficial influence on the environment, though this is otherwise possible. Note that both can easily be modeled in the current approach, either in the state transition function or in future policy function. Insofar that they are multiplicative, they should result in the same effect. Future work may tease these apart, as they may be interesting for guiding treatment.

It is worth noting that we are not the first to attempt to formalize a theory of control in MDP. \cite{HuysDayan2009} provided a first computational account of learned helplessness through simple models of action-outcome contingencies. \cite{HuysDayan2009} found that training an agent first in an unreliable environment led to later impaired learning, akin to early learned helplessness experiments, by virtue of the prior of controllability. This provides a nice demonstration of learning impairment of action-outcome contingencies. Importantly, our accounts in at least three respects. We handle sequential decision environments, which is important as it is in sequential decision environments where the consequences of pessimistic inference become most pronounced. Second, our account claims that the worst outcomes are believed, wherein their account focuses on randomness. This seems closer to the symptoms described in anxiety. Finally, our account provides insights into maintenance. With an otherwise unmodified Bayesian learner, their agents should eventually learn and be ok.

We discussed two mechanisms by which negative values can spread. There is a third that we did not address here: that of the state itself. Here we assumed that states were fully observable; in other words, there was no uncertainty about the state an agent was currently occupying. In partially-observable Markov decision problems, the state itself must be inferred from external and internal variables. This is yet another means by which bad value can spread if undue likelihood is granted to undesirable or aversive states. This ideas has previously been proposed to explain some symptoms in anxiety \citep{Paulus2012}, and not without good reason. Anxiety disorders involve negatively biased interpretation of ambiguous percepts \citep{Hartley2012}, which may result in aberrant planning and increased avoidance for similar reasons to those described above. This is an area for future research.

In this article, we have explored a computational principle of approach and avoidance learning in sequential decision environments. We have not discussed, however, the particular psychological or neurobiological mechanisms by which this sort of learning may take place. This issue has been discussed in detail elsewhere \citep{Bishop2018}, but we briefly discuss a few possibilities here. Mechanically, we understand that complex decision problems are solved by mental processes that exist on a spectrum between model-based and model-free (Daw et al., 2005). The difference between these two families of methods hinges on the reliance of an internal model of the world used to make predictions about expected long-run outcomes of particular choices.

For model-based decision making, one possibility is biased planning. In large state spaces, online sequential decision making suffers from the curve of dimensionality: in planning a series of actions to take, comparing between multiple options rapidly becomes computationally intractable when the number of options and depth of search becomes even moderately large. Thus, heuristics for reducing the size of this search can be useful. Biased pruning of search paths (such as immediately discarding paths that require some minor loss in pursuit of larger gains) may result in maladaptive decision making \citep{Huys2012}. Alternately, if planning relies on internal sampling from previously experienced episodes in order to make predictions about future value, then biased sampling may similarly result in maladaptive decision making. Recent work has shown how finite sampling during planning can result in risk aversion and the overrepresentation of low probability but strongly negative outcomes (Lieder et al., 2018). Such a mechanism is prima facie in line with results showing the availability in memory of negative outcomes being higher for patients with anxiety \citep{Borkovec1999, Miranda2007}.

Within the realm of model-free decision making, several studies have found evidence suggesting increased learning rates for negative reward prediction errors in individuals with elevated levels of anxiety \citep{Harle2017, Garrett2018}. Though an asymmetry in sensitivity to positive and negative reward prediction errors bears superficial similarity, for the reasons described above merely attending more to negative information does not necessarily predict increased avoidance insofar that avoidance can successfully isolate negative outcomes. One possibility, and topic of future research, is how prior beliefs about the reward statistics of the current environment dictate learning rates. A second possibility is prioritized replay (Russek et al., 2017; Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal mechanisms) are known to facilitate learning. It has been proposed that biased replay could bias value estimates (Gagne et al., 2018). This is another exciting area for future research.

>>>>

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/04_tree.png}}%
  }
  \par \textbf{Figure 5:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A). An optimistic agent ($w=1$) learns the optimal loss-minimizing policy through the initial large loss (B). A moderately pessimistic agent ($w=0.75$) will find both arms equally preferable insofar that it is less confident it will be able to execute optimal future choices (C). More pessimistic agents comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss (D).
\end{figure}

@@

Central to the acquisition and maintenance of anxiety disorders are symptoms including pessimistic inference, second-order conditioning, and persistent avoidance. In this article, we presented a simple computational account of these aberrations in learning and decision making. Specifically, we showed how a loss of confidence in the reliability of the future self and/or the environment can effectively backpropagate negative value across states of the environment. The result of this process are a series of inferences and behaviors resembling those observed in clinical anxiety. Though this is by no means a complete account of anxiety, the present model helps to explain the core symptoms of anxiety under one unifying framework.

Importantly, this computational account draws upon a longstanding recognition of the importance of perceived control to the development and maintenance of anxiety disorders. Central to many prominent theories of anxiety in the clinical literature highlight is a perceived lack of control. For example, learned helplessness theory and its successors claim that pathological anxiety results from a stable belief that the environment is uncontrollable, such that threat cannot be effectively mitigated. In contrast, Bandura's self-efficacy theory of anxiety posited that a lack of confidence in one's own ability to effectively handle threat was central to the pathogenesis of anxiety. Later theories, such as the triple vulnerability model, are equivocal as to whether an unreliable future environment or self is the root cause of anxiety, but highlights control beliefs as being a necessary compontent for developing anxiety. As we note above, the present model can accommodate either effect. Insofar that the state transition function and future state-action function are multiplicative, a pessimistic change to either should make similar predictions. It is possible that under certain conditions they may predict different outcomes. Future work may tease these apart, as they may be interesting for guiding treatment.

It is worth noting that we are not the first to attempt to formalize a theory of control in MDP. \cite{HuysDayan2009} provided a first computational account of learned helplessness through simple models of action-outcome contingencies. \cite{HuysDayan2009} found that training an agent first in an unreliable environment led to later impaired learning, akin to early learned helplessness experiments, by virtue of the prior of controllability. This provides a nice demonstration of learning impairment of action-outcome contingencies. Importantly, our accounts differ in at least three respects. First, we consider sequential decision environments, which is necessary for more accurately characterizing the symptoms of anxiety disorders. Second, our account specifies a mechanism by which the worst outcomes are realized; in their account, the world simply becomes random. This pessimism seems closer to what is observed in anxiety disorders. Finally, our account accounts for the persistence of these symptoms. Presumably, their model would eventually learn to escape with additional training; our account shows aberrant behaviors in the optimal (long-run) policy.

In this article, we discussed two mechanisms by which negative values can spread, through the state transition distribution and future state-action action. There is a third that we did not address here: that of the state itself. Here we assumed that states were fully observable; in other words, there was no uncertainty about the state an agent was currently occupying. In partially-observable Markov decision problems, the state itself must be inferred from external and internal variables. This is yet another means by which bad value can spread if undue likelihood is granted to undesirable or aversive states. This ideas has previously been proposed to explain some symptoms in anxiety \citep{Paulus2012}, and not without good reason. Anxiety disorders involve negatively biased interpretation of ambiguous percepts \citep{Hartley2012}, which may result in aberrant planning and increased avoidance for similar reasons to those described above. This is an area for future research.

In this article, we have explored a computational principle of approach and avoidance learning in sequential decision environments. We have not discussed, however, the particular psychological or neurobiological mechanisms by which this sort of learning may take place. This issue has been discussed in detail elsewhere \citep{Bishop2018}, but we briefly discuss a few possibilities here. One possibility is in biased sampling. If decision making and planning may rely on internal sampling from previously experienced episodes in order to make predictions about future value, then biased sampling may result in maladaptive decision making. Recent work has shown how finite sampling during planning can result in risk aversion and the overrepresentation of low probability but strongly negative outcomes (Lieder et al., 2018). Such a mechanism is prima facie in line with results showing the availability in memory of negative outcomes being higher for patients with anxiety \citep{Borkovec1999, Miranda2007}. Another possibility are an overreliance on Pavlovian heuristics, such as aversive pruning \citep{Huys2012,Lally2017}. As discussed above, future work should tease apart the predictions of our model and the aversive pruning heuristic.

Several recent studies have found evidence suggesting increased learning rates for negative reward prediction errors in individuals with elevated levels of anxiety \citep{Harle2017, Garrett2018, Aylward2019}. Though incearned learning from negative reward prediction errors may result in anxiety-like behaviors, these are likely not the sole cause of anxiety disorders for the reasons discussed above. Future research might investigate the relationship of prior beliefs about the reward statistics of the environment and learning to positive and negative prediction errors. A separate interestisng possibility is prioritized replay (Russek et al., 2017; Mattar and Daw, 2018). Offline replay of previous experiences (through hippocampal mechanisms) are known to facilitate learning. It has been proposed that biased replay could bias value estimates (Gagne et al., 2018).
